---
title:  "<br> Payrolled employments by nationality"
date: "[Ammar Ljubijankić Kutasi](mailto:ammar.ljubijankic@london.gov.uk), `r format(Sys.Date(),'%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    css:  !expr "here::here('FORMATTING','GLAstyle.css')"
    includes:
      in_header: !expr "here::here('FORMATTING','favicon.html')"
      before_body: !expr "here::here('FORMATTING','header.html')"
      after_body: !expr "here::here('FORMATTING','footer.html')"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


knitr::opts_chunk$set(echo = TRUE,scipen=999)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

```

<hr  width="100%" style="background-color: rgb(134,139,142);height: 1.0px;margin: 0;"/>

## London overview
<br/>


This release is an update of our [previous analysis on the period July 2014-June 2021](https://data.london.gov.uk/download/gla-economics-covid-19-labour-market-analysis/df64d525-e65c-4dd6-a6e5-270919aeab7e/PAYE%20Employments%20by%20nationality.html), now including the period to December 2022.

- Following the post-lockdown decrease in payrolled employments, the number of employments held by non-EU (and non-UK) workers rose steadily from January 2021 onwards while the number of EU employments stabilised. This period coincided with the end of free migration for EU citizens.

-  In November 2019, there were `r pull_data_det(dt_day="2019-11-01",nat="eu",meas="counts")` payrolled employments held by both EU and non-EU nationals living in London, each accounting for `r pull_data_det(dt_day="2019-11-01",nat="eu",meas="share",typ="%",geo="london",decimals=0)`% of all employments.

-  By December 2022, EU nationals accounted for around `r pull_data_det(dt_day="2022-11-01",nat="eu",meas="counts")` employments (`r pull_data_det(dt_day="2022-11-01", nat="eu",meas="share",typ="%",geo="london",decimals=0)`%) while non-EU nationals accounted for `r pull_data_det(dt_day="2022-11-01",nat="non_eu",meas="counts")` employments (`r pull_data_det(dt_day="2022-11-01", nat="non_eu",meas="share",typ="%",geo="london",decimals=0)`%).

- The total share of employments held by non-UK nationals has increased throughout the data period, from `r pull_data_det(dt_day="2014-07-01",nat="non_uk",meas="share",typ="%")`% in July 2014 to `r pull_data_det(dt_day="2022-12-01",nat="non_uk",meas="share",typ="%")`% in December 2022.

<foot>*Note: The count of non-EU employments include those held by workers from countries part of the EEA but not the EU. </foot>

### Total employment counts

```{r london overview count, echo= FALSE, fig.cap = figcap, out.width='100%'}

london_charts[["london_trend_counts"]]

# All charts will have the same caption, so do not copy elsewhere. If modification needed, rename appropriately.
figcap <- paste0("Source: HM Revenue and Customs – Pay As You Earn Real Time Information (non-seasonally adjusted) and Migrant Worker Scan.","<br>","<br>", "Note: Estimates are based on where employees live. Vertical lines indicates beginning of lockdowns in March 2020 and end of free migration in January 2021, respectively.")

```

<hr  width="100%" style="background-color: rgb(134,139,142);height: 1.0px;margin: 0;"/>

### Nationality shares

```{r london overview share, echo= FALSE, fig.cap = figcap, out.width='100%'}

london_charts[["london_trend_share"]]


```

<hr  width="100%" style="background-color: rgb(134,139,142);height: 1.0px;margin: 0;"/>


## Regional comparisons {.tabset}
<br/>

The UK overall has experienced a higher rate of growth in employments held by non-EU workers than London in recent years. Nevertheless, London continues to employ a much larger share of non-UK workers than the UK average.

In December 2022, `r pull_data_det(nat="eu",meas="share",typ="%",geo="london")`% of payrolled employments in London were held by EU nationals and `r pull_data_det(nat="non_eu",meas="share",typ="%",geo="london")`% were held by non-EU nationals. In both cases, this was higher than the UK average (`r pull_data_det(nat="eu",meas="share",typ="%",geo="uk")`% and `r pull_data_det(nat="non_eu",meas="share",typ="%",geo="uk")`%, respectively).

These differences are even more pronounced within certain industries, including for Hospitality and Construction:

- The share of payrolled employments held by non-UK workers in Hospitality was `r 100-as.numeric(pull_data_det(nat="uk",meas="share",typ="%",abs=FALSE,sec="Hospitality"))`% in London, more than double the UK share of `r 100-as.numeric(pull_data_det(geo="uk",nat="uk",meas="share",typ="%",abs=FALSE,sec="Hospitality"))`%.
- The share of payrolled employments held by non-UK workers in Construction was `r 100-as.numeric(pull_data_det(nat="uk",meas="share",typ="%",abs=FALSE,sec="Construction"))`% in London, four times the UK average of `r 100-as.numeric(pull_data_det(geo="uk",nat="uk",meas="share",typ="%",abs=FALSE,sec="Construction"))`%.

### Regional comparison, overall

```{r region share chart, echo= FALSE, fig.cap = figcap2, out.width='100%'}

london_charts[["emp_shares_Overall"]]

figcap2 <- paste0("Source: HM Revenue and Customs – Pay As You Earn Real Time Information (non-seasonally adjusted) and Migrant Worker Scan.","<br>","<br>", "Note: Estimates are based on where employees live.")

```

<hr  width="100%" style="background-color: rgb(134,139,142);height: 1.0px;margin: 0;"/>


### Regional comparison, Hospitality

```{r region hosp share chart, echo= FALSE, fig.cap = figcap2, out.width='100%'}

london_charts[["emp_shares_Hospitality"]]

```


<hr  width="100%" style="background-color: rgb(134,139,142);height: 1.0px;margin: 0;"/>

### Regional comparison, Construction

```{r region cons share chart, echo= FALSE, fig.cap = figcap2, out.width='100%'}

london_charts[["emp_shares_Construction"]]

```


<hr  width="100%" style="background-color: rgb(134,139,142);height: 1.0px;margin: 0;"/>

## Recent trends across industries {.tabset}
<br/>

There is variation in how payrolled employment counts have evolved across industries over the past three years, in part due to the pandemic and the end of free migration.

The charts below show trends for a selection of industries from December 2019 to December 2022. By December 2022, the number of payrolled employments in London had increased in some industries (e.g. Construction, Health) but remained well below pre-pandemic levels in others (e.g. Hospitality).

Focusing on trends by nationality, the number of payrolled employments held by EU nationals generally recovered less well over this period. There was a particularly sharp drop in EU nationals in London working in the Hospitality sector.


### Selected industries in London
```{r london facets, echo=FALSE, fig.cap = figcap, out.width='100%'}

london_charts[["fac_London"]]

```


<hr  width="100%" style="background-color: rgb(134,139,142);height: 1.0px;margin: 0;"/>

### Selected industries in the UK
```{r uk facets, echo=FALSE, fig.cap = figcap, out.width='100%'}

london_charts[["fac_UK"]]

```


<hr  width="100%" style="background-color: rgb(134,139,142);height: 1.0px;margin: 0;"/>

## Recent trends by nationality in London 

<br/>

Looking at the percentage change in payrolled employments from February 2020 to June 2021 across the ten largest industries in London, these sectoral differences become even clearer.

* Construction saw the largest rate of growth in payrolled employments at `r pull_data_det(var="p_change_feb20",nat="overall",meas="counts",typ="%",abs=TRUE,sec="Construction")`%. Over this time, the number of employment held by EU and non-EU nationals increased by `r pull_data_det(var="p_change_feb20",nat="eu",meas="counts",typ="%",abs=TRUE,sec="Construction")`% and `r pull_data_det(var="p_change_feb20",nat="non_eu",meas="counts",typ="%",abs=TRUE,sec="Construction")`%, respectively.
* Hospitality recorded the largest decrease in payrolled employments at `r pull_data_det(var="p_change_feb20",nat="overall",meas="counts",typ="%",abs=TRUE,sec="Hospitality")`%. This was driven by a particularly sharp drop in the number of employments held by EU nationals, which fell by `r pull_data_det(var="p_change_feb20",nat="eu",meas="counts",typ="%",abs=TRUE,sec="Hospitality")`%.

<br/>

::: {.infobox .info_symbol}

Note: payrolled employment counts cannot be compared with counts of total jobs. 
For the latest data on total jobs counts, see our [Labour Market Update](https://data.london.gov.uk/download/gla-economics-covid-19-labour-market-analysis/6ee07998-076c-49ff-bf8d-9f16935e31b6/LMU%20March%202022.html) release.

:::

<br/>

```{r london bar change, echo=FALSE, fig.cap = figcap3, out.width='100%'}

london_charts[["p_change_London"]]

figcap3 <- paste0("Source: HM Revenue and Customs – Pay As You Earn Real Time Information (non-seasonally adjusted) and Migrant Worker Scan.","<br>","<br>", "Note: Estimates are based on where employees live.")

```


<hr  width="100%" style="background-color: rgb(134,139,142);height: 1.0px;margin: 0;"/>

<br/>

## Table of employments
<br/>

The table below shows the number of employments by nationality in London in December 2022. The comparison month of November 2019 saw EU and non-EU employments at the same level and is therefore useful for comparing the differing trajectories in number of employments between those two groupings.

```{r london employments table, echo=FALSE, fig.cap = tabcap1, out.width='100%'}

table_list$employments_table



tabcap1 <- paste0("Source: HM Revenue and Customs – Pay As You Earn Real Time Information (non-seasonally adjusted) and Migrant Worker Scan.","<br>","<br>", "Note: Estimates are based on where employees live. Comparison is to November 2019 when the number of employments with EU and non-EU nationals were the same.")
                  
```

<hr  width="100%" style="background-color: rgb(134,139,142);height: 1.0px;margin: 0;"/>

## Understanding the data
<br/>

HM Revenue and Customs (HMRC) has published statistics on [payrolled employments in the UK by region, industry and nationality](https://www.gov.uk/government/statistics/uk-payrolled-employments-by-nationality-region-and-industry-from-july-2014-to-december-2022), including data on the period from June 2021 to December 2022. This article explores some of the key trends for London.

Please note, the payrolled employment counts used in this article cannot be:

-  directly compared with payrolled employee numbers - payrolled employees can have multiple payrolled employments, so employment counts follow the same trend as employee counts, but will be higher.

-  directly compared with total jobs numbers - these statistics only include employments from the PAYE system, they do not include employments from non-payrolled sources such as self-employment. 

-  used as a count of monthly or annual migration to the UK or London, or to show the number of non-UK nationals who are currently living in the UK or London.

Nationality breakdowns are determined using HMRC’s Migrant Worker Scan (MWS). Nationality is reported by individuals when they register for a National Insurance number and may change following registration. 

The data has not been seasonally adjusted.

<br/>
<hr  width="100%" style="background-color: rgb(134,139,142);height: 1.0px;margin: 0;"/>

## Key dates
<br/>

This release concerns the movement of workers from abroad, including the EU, to the UK. In this context, the following dates are relevant, with most related to the process of the UK leaving the EU.

- 23 June 2016 - the UK held a referendum and 52% of those who voted chose to leave the EU.

- 31 January 2020 - the UK left the EU and entered a transition period.

- 23 March 2020 - the first COVID-19 lockdown was announced and the Coronavirus Job Retention Scheme was introduced.

- 31 December 2020 - the transition period ended and the UK left the EU single market and customs union. Free movement between the EU and the UK ended.

- 1 January 2021 - the UK introduced a new, points-based immigration system.

<br/>
<hr  width="100%" style="background-color: rgb(134,139,142);height: 1.0px;margin: 0;"/>

::: {.tocify-extend-page data-unique="tocify-extend-page" style="height: 0;"}
:::